Variation in gaze understanding across the life span: A process-level perspective

### 2023-10-24 paper/manuscript.Rmd
ready for first submission & pre-print

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship"

abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze-following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze understanding across the lifespan (*N* = 471 3- to 80-years-olds; children from a mid-sized German city; international remotely tested adults). We found a steep learning curve during preschool years, in which children become more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze understanding in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we tested two implications of our gaze model, namely that gaze understanding should be related to perspective-taking abilities and vector estimation in non-social settings (*N* = 102 4- to 5-year-olds). We found that gaze understanding is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "7909"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

Our gaze model assumed vector calculations on a process level. By isolating physical vector estimation experimentally, we could show that gaze understanding does indeed, to a certain degree, rely on this component.  However, physical vector estimation alone did not suffice to explain variation in gaze understanding. Additionally, perspective-taking proved to be a relevant social-cognitive ability when predicting the performance in the TANGO. The overall ToM aggregate score did not add explanatory power.

The TANGO and the two perspective-taking tasks can be seen as instances of visuospatial perspective-taking. Often, researchers distinguish between Level 1 and Level 2 perspective-taking tasks. Level 1 perspective-taking is concerned with the visibility of objects from a particular viewpoint (Surtees et al., 2013), and can usually be solved independently from another’s agent frame of reference and without mental rotation [not matching the “mentalizing criterion”; @quesque2020theoryofmind]. In contrast, Level 2 perspective-taking tasks ask participants to judge how (visually or conceptually) the world looks for another person. Solving these tasks presumably relies on a simulation of mentally transforming one's own body schema into the space of another agent [@erle2017grounded]. Participants must understand that different viewpoints lead to different perspectives [@flavell1981younga]: two agents can see the same object "in different, incompatible, fine-grained ways" [@rakoczy2022foundations]. In short, Level 1 perspective-taking addresses *what* an agent can see, while Level 2 perspective-taking addresses *how or where* they see it [@moll2006level]. The TANGO falls into the first category - however, in contrast to existing research, on a continuous scale - while the perspective-taking tasks within our ToM scale fall into the second category.

@@

In this paper, we have focused on studying variation across ages and individuals. However, our findings rely on participants from Western, Educated, Industrialized, Rich, and Democratic (WEIRD) backgrounds [@henrich2010weirdesta]. Cultural variation has been found in many foundational aspects of cognition and socialization: for example, in parent-child interaction and communicative signals [@nielsen2017persistent]. First evidence suggests that cultural variation in face-to-face interactions does not influence infants' gaze-following abilities [@hernik2019infanta]. While we cannot generalize our here-reported developmental trajectory to different socio-environmental settings, we predict that our presented process-model of gaze understanding holds true across communities. Analyzing cross-cultural variation and checking for predicted "signature patterns" in the data will inform our modeling work and theory building in further detail [@amir2020crosscultural].

>>>>

affiliation:
  - id            : "1"
    institution   : "Department of Comparative Cultural Psychology, Max Planck Institute for Evolutionary Anthropology, Leipzig, Germany"
  - id            : "2"
    institution   : "Institute of Psychology, Leuphana University Lüneburg, Germany"
  - id            : "3"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "*"
    institution   : "Shared senior authorship" 
abstract: |
  Following eye gaze is fundamental for many social-cognitive abilities, for example, when judging what another agent can or cannot know. While the emergence of gaze-following has been thoroughly studied on a group level, we know little about (a) the developmental trajectory beyond infancy and (b) the sources of individual differences. In Study 1, we examined gaze understanding across the lifespan (*N* = 471 3- to 80-years-olds; children from a mid-sized German city; international remotely tested adults). We found a steep learning curve during preschool years, in which children became more precise in locating the attentional focus of an agent. Precision levels then stayed comparably stable throughout adulthood with a minor decline toward old age. In Study 2, we formalized the process of gaze understanding in a computational cognitive model that allowed us to conceptualize individual differences in a psychologically meaningful way (*N* = 60 3- to 5-year-olds, 50 adults). According to our model, participants estimate pupil angles with varying levels of precision based on observing the pupil location within the agent's eyes. In Study 3, we tested two implications of our gaze model, namely that gaze understanding could be related to vector estimation in non-social settings and perspective-taking abilities (*N* = 102 4- to 5-year-olds). We found that gaze understanding is associated with both of these abilities but less so with other Theory of Mind tasks. This work illustrates how the combination of reliable measurement instruments and formal theoretical models allows us to explore the in(ter)dependence of core social-cognitive processes in greater detail.
keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "7914"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

Our gaze model assumed vector calculations on a process level. By isolating physical vector estimation experimentally, we could show that gaze understanding does indeed, to a certain degree, rely on this component. However, physical vector estimation alone did not suffice to explain variation in gaze understanding. Additionally, perspective-taking proved to be a relevant social-cognitive ability when predicting the performance in the TANGO [cf. @astor2022gaze]. The overall ToM aggregate score did not add explanatory power.

The TANGO and the two perspective-taking tasks can be seen as instances of visuospatial perspective-taking. Often, researchers distinguish between Level 1 and Level 2 perspective-taking tasks. Level 1 perspective-taking is concerned with the visibility of objects from a particular viewpoint [@surtees2013similarities], and can usually be solved independently from another’s agent frame of reference and without mental rotation [not matching the “mentalizing criterion”; @quesque2020theoryofmind]. In contrast, Level 2 perspective-taking tasks ask participants to judge how (visually or conceptually) the world looks for another person. Solving these tasks presumably relies on a simulation of mentally transforming one's own body schema into the space of another agent [@erle2017grounded]. Participants must understand that different viewpoints lead to different perspectives [@flavell1981younga]: two agents can see the same object "in different, incompatible, fine-grained ways" [@rakoczy2022foundations]. In short, Level 1 perspective-taking addresses *what* an agent can see, while Level 2 perspective-taking addresses *how or where* they see it [@moll2006level]. The TANGO falls into the first category - however, in contrast to existing research, on a continuous scale - while the perspective-taking tasks within our ToM scale fall into the second category.

@@

In this paper, we have focused on studying variation across ages and individuals. However, our findings rely on participants from Western, Educated, Industrialized, Rich, and Democratic (WEIRD) backgrounds [@henrich2010weirdesta]. Cultural variation has been found in many foundational aspects of cognition and socialization: for example, in parent-child interaction and communicative signals [@nielsen2017persistent]. First evidence suggests that cultural variation in face-to-face interactions does not influence infants' gaze shifts [@hernik2019infanta]. While we cannot generalize our here-reported developmental trajectory to different socio-environmental settings, we predict that our presented process-model of gaze understanding holds true across communities. Analyzing cross-cultural variation and checking for predicted "signature patterns" in the data will inform our modeling work and theory building in further detail [@amir2020crosscultural].
